---
title: "Multiple regression of microbiome and IBD-related variables using HMP datasets"
author: "David Xin Zhao"
date: "Last edited `r format(Sys.time(), '%d %B %Y')`"
knit: (function(inputFile, encoding) { 
      out_dir <- 'docs';
      rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_file=file.path(dirname(inputFile), out_dir, 'index.html'))}) 
output:
 html_document:
  theme: 
   bootswatch: simplex
    bg: "#202121"
    fg: "#B8BCC2" 
  # highlight: tango
    # css: styles.css 
  toc: TRUE
  toc_float: TRUE
     # collapsed: false
     # smooth_scroll: false 
  # toc_depth: 2
  number_sections: TRUE
  fig_width: 7
  fig_height: 6
  fig_caption: TRUE
  df_print: paged
---

```{=html}
<style>
#TOC{
 color: #7E7C73;
 font-family: Calibri;
 font-size: 16px;
 border-color: #BBC4C2;
}
h1.title{
 color: #140005;
 background-color: #F5F5F5;
 opacity: 0.6;
 font-family: Calibri;
 font-size: 30px;
}
h4.author{
 color: #464033;
 font-family: Calibri;
 font-size: 25px;
}
h4.date{
 color: #464033;
 font-family:Calibri; 
}
body{
 color: #464033; 
 font-family: Calibri; 
 background-color: #F5F5F5;
}
pre{
 color: #464033;
 background-color: #FFFFFF;
}
</style>
```

```{r setup, include = FALSE}
# set options for the entire document 
knitr::opts_chunk$set(fig.align = 'center', 
                      fig.dim =c(6,4), 
                      dev="png",
                      echo=TRUE, #display code in output document 
                      error=FALSE,
                      message=FALSE) #stop render when error occurs   
```


This project was built with R `r getRversion()`.
```{r load libraries, message=FALSE, warning=FALSE}

# libraries loaded 
# if(!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("Maaslin2")

library(Maaslin2) # implement multivariate regression 
library(tidyverse)
library(knitr)
library(janitor) # create descripitive stats table 
library(rmarkdown) # create paged tables 
library(ggrepel) # organize labels nicely 
library(ggplot2)
library(bslib) # bootswatch themes 

# use the function below to list available html output themes 
# bslib::bootswatch_themes()
# ?bs_theme to see more variables adjustable in html themes 

```

# Data preprocessing
## Raw datasets 

Set the folder location for the microbiome abundance data set 
```{r}
input_data <- system.file("extdata",
                          "HMP2_taxonomy.tsv",
                          package="Maaslin2") # the abundance table file 

```

Set the folder location for the meta data set 
```{r metadata location}
input_metadata <- system.file("extdata",
                              "HMP2_metadata.tsv",
                              package = "Maaslin2")  # the metadata table file 

```

Read in the microbiome abundance data 
```{r microbiom data, collapse=TRUE}

df_input_data <- read.table(file = input_data,
                            header = TRUE,
                            sep = "\t",
                            row.names = 1,
                            stringsAsFactors = FALSE) 

head(df_input_data, 5) # view first rows 
```

Read in the metadata 
```{r metadata}

df_input_metadata <- read.table(file=input_metadata,
                                header = T,
                                sep="\t",
                                row.names = 1,
                                stringsAsFactors = F) 

head(df_input_metadata, 5) # view first rows 
```


## Exploratory data analysis
Generate descriptive statistics for microbiome relative abundance data, calculating mean, standard deviations and rate of missing values 
```{r EDA-microbiome abundance data}
# descriptive stats for top 20 abundant genus 
df_input_data_long <- df_input_data %>%
  rownames_to_column("sample") %>%
  gather(key="genus", value="rel_abund", -sample) #reshape short to long data frame 

summary_genus <- df_input_data_long %>% 
  group_by(genus) %>%
  summarise(average = mean(rel_abund, na.rm=TRUE)*100,
            std = sd(rel_abund, na.rm=TRUE)*100,
            zero_rate = mean(rel_abund == 0)*100)%>%
  arrange(desc(average)) %>%
  head(20)  # top 20 genus based on average relative abundance 


# display summary of relative abundance at genus level in a table 
kable(summary_genus, 
      col.names = c("Species",
                    "Average percentage",
                    "Standard deviation",
                    "Zeroness percentage"),
      caption = 'Descriptive statistics of top 20 abundant species',
      format = 'html',
      digits = 2) 

```


```{r}
# inspect metadata 

df_input_metadata # print paged table 

```


Visualize relationship between representative genus and dysbiosis (continuous variable), stratified by diagnosis status
```{r scatter plot-microbiome vs dysbiosis}

#extract top 10 genus with highest average of relative abundance 
top10_g <- summary_genus %>% 
  arrange(desc(average)) %>% 
  head(10) %>% 
  pull(genus)

#select top 10 representative genus 
df_input_data_top <- df_input_data %>% 
  select(all_of(top10_g)) %>% 
  rownames_to_column("subject_id")

#subset dysbiosis and diagnosis variables 
df_input_metadata_sub <- df_input_metadata %>% 
  select(diagnosis, dysbiosis) %>% 
  rownames_to_column("subject_id") 

#merge microbiome data and metadata 
merge_micro_meta_top <- df_input_data_top %>% 
  inner_join(df_input_metadata_sub, by="subject_id")

```


```{r plot, message=FALSE, fig.cap='Relationship between representative top genus and diagnosis'}
# scatter plot given genus vs dysbiosis with faceting by diagnosis status (for instance, F. prausnitzii)
require(RColorBrewer)

ggplot(merge_micro_meta_top, aes(x= dysbiosis, 
                                 y=Faecalibacterium.prausnitzii,
                                 color=diagnosis))+
  geom_point(alpha=0.7)+
  geom_smooth(method = "lm",
              se=FALSE)+
  scale_color_brewer(palette = "Dark2")+
  theme_minimal()
  

```



There were 11 covariates in mete-data as follows, 
`age`, `dysbiosis`, and `collection` were numerical variables; and `site`, `diagnosis`, `subject`, `antibiotics`, `dysbiosis_binary`, `dysbiosisnonIBD`, `dysbiosisUC`, `dysbiosisCD` were character variables. 

In addition,  missing values were present in three variables, `dysbiosisnonIBD`, `dysbiosisUC`, `dysbosisCD`. 

```{r convert to factor variables}

# convert some character variables to factor variables 
# diagnosis 
df_input_metadata$diagnosis <- factor(df_input_metadata$diagnosis, 
                                      levels = c("nonIBD", # set "nonIBD" a reference level 
                                                 "UC",
                                                 "CD"))

# convert remaining character variables to binary factor variables  
df_input_metadata <- df_input_metadata %>% 
  mutate(site = factor(site),
         antibiotics = factor(antibiotics),  
         dysbiosis_binary = factor(dysbiosis_binary),
         dysbiosisnonIBD = factor(dysbiosisnonIBD),
         dysbiosisUC = factor(dysbiosisUC),
         dysbiosisCD = factor(dysbiosisCD)) 

# summary(df_input_metadata) # investigate again the resulting meta-data 

```

Calculate median and IQR for numerical variables; and subtotal (percentages) for categorical variables. Conduct chi-square test on categorical variables and do ANOVA on numerical variables within each strata according to diagnosis.Use a table to display summary. 
```{r summary table for meta-data}

# site distribution by diagnosis levels 
df_site <- tabyl(df_input_metadata, site, diagnosis) %>% 
  adorn_totals(c("row", "col"))%>%
  adorn_percentages("row") %>% 
  adorn_pct_formatting(digits = 0) %>% 
  adorn_ns() %>%  # combine count and percentage 
  add_row(site = "site", 
          .before = 1) %>%  #add one row showing variable name 
  filter(row_number() !=7) #remove total row 
```

Follow the instruction in [programming with dyplyr](<https://dplyr.tidyverse.org/articles/programming.html>)
```{r build a function, echo=FALSE} 

#write a function to generate a summary table for covariates 
count_pct <- function(x){
  data1 <- df_input_metadata %>% 
    tabyl(., .data[[x]], diagnosis) %>% #follow the website, "programming with dplyr"  
    adorn_totals(c("row", "col"))%>%
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits = 0) %>% 
    adorn_ns()
  
  addrow <- data.frame(colnames(data1)[1], NA, NA, NA, NA)
  
  names(addrow) <- colnames(data1)
  
  data2 <- rbind(addrow, data1) 
  
  names(data2)[1] <- "covariable" 
  
  data2 %>%
    filter(row_number() != n())
} 

```

Apply the function to multiple categorical variables. Subset only categorical variables from `df_input_metadata`. 
```{r}
cat_cov <- df_input_metadata %>% 
  select(where(is.factor)) %>% 
  select(-diagnosis) %>% 
  colnames()

sum_cat <- lapply(cat_cov, count_pct) #list of summary of multiple categorical variables  

# combine each element of the list by rows before storing in a data frame 
df_sum_cat <- as.data.frame(do.call(rbind, sum_cat))

```


```{r, results='hide'}

str(df_input_metadata)

numerical_summary <- function(x){
dat_ns <- df_input_metadata %>% 
  group_by(diagnosis) %>% 
  summarise(average = round(mean(.data[[x]], na.rm=T),2),
            std = round(sd(.data[[x]], na.rm = T),2)) %>% 
  select(-average)
  
df_num_sub <- df_input_metadata %>% 
  group_by(diagnosis) %>% 
  summarise(average = round(mean(.data[[x]], na.rm=T),2)) %>%
  adorn_ns(ns = dat_ns)%>% 
  rename("average (sd)"= average) %>%
  spread(key=diagnosis, value = "average (sd)") %>% 
  mutate(covariable = x) %>% 
  select(covariable, 1:3)


total_dat_ns <- df_input_metadata %>%
  summarise(average = round(mean(.data[[x]], na.rm=T),2),
            std = round(sd(.data[[x]], na.rm = T),2)) %>%
  select(-average) %>%
  mutate(covariable = x) %>% 
  select(covariable, std)
  
df_num <- df_input_metadata %>%
  summarise(average = round(mean(.data[[x]], na.rm=T),2)) %>%
  mutate(covariable = x) %>% 
  select(covariable, average) %>% 
  adorn_ns(ns = total_dat_ns) %>% 
  rename("average (sd)" = average)

df_num_bind <- cbind(df_num_sub, df_num[,2]) 

colnames(df_num_bind)[5] <- "overall"

print(df_num_bind) 
}

numerical_summary("age") #the resulting function works properly

```

Apply the function `numerical_summary` to all the numerical variables in meta-data with `lapply`.  
```{r, collapse=TRUE, results='hide'} 
# extract column names for all the numerical variables in meta-data 
num_names <- df_input_metadata %>% 
  select(where(is.numeric)) %>% 
  colnames()

list_num_summary <- lapply(num_names, numerical_summary) 
  
# str(list_num_summary) 

# convert individual element of the list to data frames and bind them into a single data frame 
df_num_summary <- as.data.frame(do.call(rbind, list_num_summary)) 

```

Combine descriptive stats tables, `df_sum_cat` and `df_num_summary` for categorical and numerical covariables, respectively  
```{r}
# head(df_sum_cat, 3)
# head(df_num_summary, 3)

names(df_sum_cat)[5] <- "overall"  #rename the column name

df_all_cov <- bind_rows(df_sum_cat, df_num_summary) #bind by rows 

#replace NA in the column, "co variable" with " unknown"
df_all_cov$covariable <- as.character(df_all_cov$covariable)
df_all_cov$covariable %>% replace_na("unknown")  

```

Apply chi-square test and ANOVA to categorical descriptive stats and numerical descriptive stats, respectively 
```{r}

# write a function to do chi-square to categorical variables 
chi_square <- function(x){
  chisq <- chisq.test(df_input_metadata[[x]], df_input_metadata$diagnosis)
  
  pval <- chisq$p.value
  
  pval_digit <- format(round(pval, 3), nsmall=4)
  
  print(pval_digit)
  
}

# chi_square("site")

df_pval_cat <- data.frame(covariable = cat_cov, 
                          pvalue = numeric(length = length(cat_cov)))   #create empty df to store p-values 

df_pval_cat$pvalue <- sapply(cat_cov, chi_square) 


#write a function to do anova to numerical variables against diagnosis 
anova_num <- function(x){
  f <- paste0(x, "~", "diagnosis")
  
  pval <- anova(lm(f, data=df_input_metadata))$"Pr(>F)"[1]
  
  pval_digit <- format(round(pval, 3), nsmall=5)
  
  print(pval_digit)
}

df_pval_num <- data.frame(covariable = num_names, 
                          pvalue = numeric(length = length(num_names)))   #create empty df to store p-values 

df_pval_num$pvalue <- sapply(num_names, anova_num) 


#join p-values with descriptive stats table 
df_all_cov_pval <- df_all_cov %>% 
  left_join(df_pval_num, by="covariable") %>% 
  left_join(df_pval_cat, by="covariable")  

df_all_cov_pval$pvalue.x <- if_else(!is.na(df_all_cov_pval$pvalue.x), 
                            df_all_cov_pval$pvalue.x,
                            df_all_cov_pval$pvalue.y)  

df_all_cov_pval$covariable <- if_else(!is.na(df_all_cov_pval$covariable), 
                            df_all_cov_pval$covariable,
                            "unknown")  

df_all_cov_pval2 <- df_all_cov_pval %>% select(-pvalue.y) %>% rename(pvalue = pvalue.x) 

df_all_cov_pval2 #view the resulting table

```

# Fit the mdoel
Run MaAsLin2 on the HMP2 data, building a multiple regression model to evaluate the association between microbial species (relative abundance) and `dysbiosis` scores adjusting for IBD `diagnosis` status.

Key arguments were set as followings,\  
- linear regression was used for data modeling `analysis_method` = `LM` 
- set the minimal threshold for taxa prevalence as 0.1 `min_prevalence` = `0.1`
- set the minimal threshold for taxa relative abundance as 0.01% `min_abundance` = `0.0001`
- fixed effect variables `fixed_effects` = `c("diagnosis", "dysbiosis")`
- random effect variable `random_effects` = `c("subject")`  
- set the baseline for the categorical variable `diagnosis` `reference` = `c("diagnosis","nonIBD")`

Note that MaAsLin2 can not yet handle interaction terms directly in model building. Alternatively, persons can add additional interactive covariables in the input data frame prior to modeling. 

```{r message=FALSE,results='hide', collapse=TRUE}
fit_data2 <- Maaslin2(
        input_data = df_input_data,
        input_metadata = df_input_metadata,
        min_prevalence = 0.1, #typical prevalence threshold: 10-50% (up to 70-90% can be reasonable)
        min_abundance = 0.0001, #typical relative abundance threshold: 0.0001 
        normalization = "NONE",  #without additional normalization  
        output = "demo_output2",
        max_significance = 0.1,
        plot_heatmap = FALSE,
        plot_scatter = FALSE,
        analysis_method = "LM",  #linear regression model 
        fixed_effects = c("diagnosis", "dysbiosis"), # fixed effect variables, diagnosis and dysbiosis 
        random_effects = c("subject"), # random effect variable  
        reference = c("diagnosis,nonIBD")) #set the baseline for the categorical variable, diagnosis 

all_results <- fit_data2$results #extract coefficients and q-value

```

## Significant taxa checkup
significant associations is provided in `significant_results.tsv`

```{r}
sig_results <- all_results %>% 
        filter(qval <= 0.05) %>% #sig threshold q value = 0.05 
        arrange(qval)

sig_results  # print in the paged table format 

```

Draw a volcano plot to illustrate q-values and coefficients 
```{r volcano plot qval and coeff, warning=FALSE, include=TRUE, fig.cap='Illustration of coefficients and q-values of diagnosis scores for individual species'}
# add a column to indicate significant genus 
all_results_sig <- all_results %>% 
  mutate(whether_sig = case_when(coef >= 0.3 & qval <= 0.05 ~ "Positive",
                                 coef <= -0.3 & qval <= 0.05 ~ "Negative",
                                 abs(coef) < 0.3 | qval >0.05 ~ "No association")) 


# add label column 
all_results_sig$relabel <- NA

all_results_sig$relabel[all_results_sig$whether_sig != "No association"] <- all_results_sig$feature[all_results_sig$whether_sig != "No association"]

# set personal color palate 
mycolors <- c("blue", "red", "black")
names(mycolors) <- c("Negative", "Positive", "No association") 

# volcano plot 
all_results_sig %>% 
  filter(value == "dysbiosis") %>% #focus on dysbiosis 
ggplot(aes(x=coef, y=-log10(qval),
           color=whether_sig,
           label=relabel))+
  geom_point()+
  geom_vline(xintercept = c(-0.3, 0.3), color="red")+
  geom_hline(yintercept = -log10(0.05), color="red")+
  scale_color_manual(values = mycolors)+
  theme_minimal()+ 
  geom_text_repel() #label taxa names 

```

## Model interpretation
Use `Faecalibacterium.prausnitzii` as an example for model interpretation. 
```{r}
coef_Fae <- all_results_sig %>% 
  filter(value == "dysbiosis") %>%
  filter(feature == "Faecalibacterium.prausnitzii") %>%
  select(coef) %>% 
  pull() %>%
  round(., 3) 

qval_Fae <- all_results_sig %>% 
  filter(value == "dysbiosis") %>%
  filter(feature == "Faecalibacterium.prausnitzii") %>%
  select(qval) %>% 
  pull() %>%
  format(., scientific=TRUE) 
```

*F. prausnitzii* is negatively associated with `dysbiosis` (q value = `r qval_Fae`). In particularly, *F. prausnitzii* percentage in feces may decrease by `r abs(coef_Fae)` on average when `dysbiosis` scores increases by 1 unit independent of `diagnosis` status. 

## Session information
```{r package version info}
sessionInfo()  
```


















