---
title: "Multivariable regression of microbiome and IBD using MaAsLin2 R package"
author: "David Xin Zhao"
date: "Last edited `r format(Sys.time(), '%d %B %Y')`"
knit: (function(inputFile, encoding) { 
      out_dir <- 'docs';
      rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_file=file.path(dirname(inputFile), out_dir, 'maaslin2-IBD-microbiome.html')) })
output:
  html_document:
    theme: paper
    # css: styles.css
    df_print: paged 
    toc: yes
    toc_float: yes
    collapsed: false 
  toc_depth: 3
  number_sections: true
---


```{=html}
<style>
#TOC{
 color: #7E7C73;
 font-family: Calibri;
 font-size: 16px;
 border-color: #BBC4C2;
}
h1.title{
 color: #140005;
 background-color: #F5F5F5;
 opacity: 0.6;
 font-family: Calibri;
 font-size: 30px;
}
h4.author{
 color: #464033;
 font-family: Calibri;
 font-size: 25px;
}
h4.date{
 color: #464033;
 font-family:Calibri; 
}
body{
 color: #464033; 
 font-family: Calibri; 
 background-color: #F5F5F5;
}
pre{
 color: #464033;
 background-color: #FFFFFF;
}
</style>
```


```{r setup, include = FALSE}
# set options for the entire document 
knitr::opts_chunk$set(fig.align = 'center', 
                      echo=TRUE, #display code in output document 
                      error=FALSE,
                      message=FALSE,
                      dev="png",
                      fig.dim=c(6,4))   
```


## Install MaAsLin2 with Bioconductor

```{r message=FALSE}

# if(!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("Maaslin2")

library(Maaslin2)
library(tidyverse)

```

## Multiple regression of microbiome and clincial covariates of interest

### Load data sets (use example datasets provided by MaAsLin2 package)

```{r include=FALSE}

# set file pathway to abundance data 
input_data <- system.file("extdata",
                          "HMP2_taxonomy.tsv",
                          package="Maaslin2") # the abundance table file 

# set file pathway to meta data 
input_metadata <- system.file("extdata",
                              "HMP2_metadata.tsv",
                              package = "Maaslin2")  # the metadata table file 

# # get the pathway (functional) data 
# download.file("https://raw.githubusercontent.com/biobakery/biobakery_workflows/master/examples/tutorial/stats_vis/input/pathabundance_relab.tsv", "./pathabundance_relab.tsv")

```

```{r message=FALSE, results='hide', collapse=TRUE}
 
# relative abundance data 
df_input_data <- read.table(file = input_data,
                            header = TRUE,
                            sep = "\t",
                            row.names = 1,
                            stringsAsFactors = FALSE) 

# meta data 
df_input_metadata <- read.table(file=input_metadata,
                                header = T,
                                sep="\t",
                                row.names = 1,
                                stringsAsFactors = F) 

# pathway data 
df_input_path <- data.frame(read.csv("./pathabundance_relab.tsv", 
                                     sep = "\t", 
                                     stringsAsFactors = F, 
                                     row.names = 1))

```

### Exploratory data analysis
Generate descriptive statistics for microbiome relative abundance data, calculating mean, standard deviations and rate of missing values 
```{r EDA-microbiome abundance data, echo=FALSE}
# descriptive stats for top 20 abundant genus 
df_input_data_long <- df_input_data %>%
  rownames_to_column("sample") %>%
  gather(key="genus", value="rel_abund", -sample) #reshape short to long data frame 

summary_genus <- df_input_data_long %>% 
  group_by(genus) %>%
  summarise(average = mean(rel_abund, na.rm=TRUE)*100,
            std = sd(rel_abund, na.rm=TRUE)*100,
            zero_rate = mean(rel_abund == 0)*100)%>%
  arrange(desc(average)) %>%
  head(20)  # top 20 genus based on average relative abundance 


# display summary of relative abundance at genus level in a table 
knitr::kable(summary_genus, 
      col.names = c("Genus",
                    "Average percentage",
                    "Standard deviation",
                    "Zeroness percentage")) 

```


```{r}
# structure of metadata 
str(df_input_metadata)

df_input_metadata # print paged table 

```


Visualize relationship between representative genus and dysbiosis (continuous variable), stratified by diagnosis status
```{r scatter plot-microbiome vs dysbiosis}

#extract top 10 genus with highest average of relative abundance 
top10_g <- summary_genus %>% 
  arrange(desc(average)) %>% 
  head(10) %>% 
  pull(genus)

#select top 10 representative genus 
df_input_data_top <- df_input_data %>% 
  select(all_of(top10_g)) %>% 
  rownames_to_column("subject_id")

#subset dysbiosis and diagnosis variables 
df_input_metadata_sub <- df_input_metadata %>% 
  select(diagnosis, dysbiosis) %>% 
  rownames_to_column("subject_id") 

#merge microbiome data and metadata 
merge_micro_meta_top <- df_input_data_top %>% 
  inner_join(df_input_metadata_sub, by="subject_id")

```


```{r plot, message=FALSE}
# scatter plot given genus vs dysbiosis with faceting by diagnosis status (for instance, F. prausnitzii)
require(RColorBrewer)

ggplot(merge_micro_meta_top, aes(x= dysbiosis, 
                                 y=Faecalibacterium.prausnitzii,
                                 color=diagnosis))+
  geom_point(alpha=0.7)+
  geom_smooth(method = "lm",
              se=FALSE)+
  scale_color_brewer(palette = "Dark2")+
  theme_minimal()
  

```


```{r}
# descriptive stats of pathway data set 
transp_input_path <- data.frame(t(df_input_path)) # transpose df 

transp_input_path_long <- transp_input_path %>%
  rownames_to_column("sample") %>%
  gather(key="pathway", value="rel_abund", -sample) # reshape to a long format 


summary_path <- transp_input_path_long %>% 
  group_by(pathway) %>%
  summarise(average = mean(rel_abund, na.rm=TRUE),
            std = sd(rel_abund, na.rm=TRUE),
            zero_rate = mean(rel_abund == 0)*100)%>%
  arrange(desc(average)) %>%
  head(20)  # top 20 pathways based on average relative abundance 


knitr::kable(summary_path, 
      col.names = c("Pathway",
                    "Average",
                    "Standard deviation",
                    "Zeroness percentage")) 

```





### Fit mutiple linear regression model on microbiome data

The following command runs MaAsLin2 on the HMP2 data, running a multivariable regression model to test for the association between microbial species abundance versus IBD diagnosis and dysbiosis scores.

```{r message=FALSE,results='hide',collapse=TRUE}
fit_data2 <- Maaslin2(
        input_data = df_input_data,
        input_metadata = df_input_metadata,
        min_prevalence = 0, #without additional filtering 
        normalization = "NONE",  #without additional normalization  
        output = "demo_output2",
        max_significance = 0.1,
        plot_heatmap = FALSE,
        plot_scatter = FALSE,
        analysis_method = "LM",  #linear regression model 
        fixed_effects = c("diagnosis", "dysbiosis"),  #set diagnosis and dysbiosis as fixed effect variables 
        reference = c("diagnosis,nonIBD")) #set the baseline for the categorical variable, diagnosis 

all_results <- fit_data2$results #extract coefficients and q-value

```

### Evaluate significant genus based on model parameters

significant associations is provided in 'significant_results.tsv'

```{r}
sig_results <- all_results %>% 
        filter(qval <= 0.05) %>% #sig threshold 
        arrange(qval)

head(sig_results, 5) #view first rows 

```

### Fit multivariable regression models on functional data

Authors tend to work with CPM units because we find them to be more convenient, but they are numerically equivalent to relative abundances for modeling purposes (CPM = RA \* 1e6)

Quick tips for using MaAsLin2 on functional profiles: 1.To reduce the number of features, you typically want to run either the unstratified functional features ( or a filtered subset of the stratified features) 2.You may also want to remove functional features that are highly correlated to one specific taxon (i.e. likely contributed by that microbe), since these can be better-explained by taxonomic changes

```{r normalization, message=FALSE}

#This can also be done with with the HUMAnN 3 untiliy `humann_split_stratified_table`
unstrat_pathways <-function(dat_path){
  temp = dat_path[!grepl("\\|",rownames(dat_path)),]
  return(temp)
}

df_input_path = unstrat_pathways(df_input_path)

```

Run the same model as above on the MetaCyc pathway table generated from the bioBakery workflows.

```{r message=FALSE,results='hide',collapse=TRUE}
# fit the model on pathway (functional) data - it takes a few minutes  
fit_func <- Maaslin2(
        input_data = df_input_path,
        input_metadata = df_input_metadata,
        output = "demo_functional",
        fixed_effects = c("diagnosis", "dysbiosis"),
        reference=c("diagnosis,nonIBD"),
        min_abundance = 0.0001,
        min_prevalence = 0.1) 

# note: here we have the samples in the columns and the features in the rows and MaAsLin2 correctly identified this and was able to match the samples. 

all_results_func <- fit_func$results

```

### Interactions

Unfortunately, MaAsLin2 does not yet provide a direct interface for testing for interaction. Instead, the user needs to create artificial interaction columns as additional 'fixed_effects' terms.

Using the above fit as an example, to test for the interaction between 'diagnosis' and 'dysbiosis', I can create two additional columns: 'CD_dysbiosis' and 'UC_dysbiosis' (since the reference for 'diagnosis' is 'nonIBD'):

```{r interaction term, message=FALSE,results='hide',collapse=TRUE}

df_input_metadata$CD_dysbiosis <- (df_input_metadata$diagnosis == "CD") * df_input_metadata$dysbiosis

df_input_metadata$UC_dysbiosis <- (df_input_metadata$diagnosis == "UC") * df_input_metadata$dysbiosis

fit_data_interaction <- Maaslin2(
        input_data = df_input_data,
        input_metadata = df_input_metadata,
        min_prevalence = 0,
        normalization = "NONE",
        output = "demo_output_interaction",
        fixed_effects = c("diagnosis", "dysbiosis", "CD_dysbiosis", "UC_dysbiosis"),
        reference = c("diagnosis,nonIBD")) 

all_results_interaction <- fit_data_interaction$results  # store results 

```

### Random effects

We note that HMP2 is a longitudinal design where the same subject (column 'subject') can have multiple samples. We thus ask MaAsLin2 to use subject as its random effect grouping variable:

```{r mixed effect model, message=FALSE,results='hide',collapse=TRUE}

fit_data_random <- Maaslin2(
        input_data = df_input_data,
        input_metadata = df_input_metadata,
        min_prevalence = 0,
        normalization = "NONE",
        output = "demo_output_random",
        fixed_effects = c("diagnosis", "dysbiosis"),
        random_effects = c("subject"),
        reference = c("diagnosis,nonIBD")
)

all_results_random <- fit_data_random$results

```

If you are interested in testing the effect of time in a longitudinal study, then the itme point variable should be included in 'fixed_effects' during your MaAsLin2 call.

### Prevalence and abundance filtering in MaAsLin2

minimum prevalence threshold: 10-50% (up to 70-90% can be reasonable) minimum relative abundance threshold: 0.0001

```{r prevalence and abundance filtering, message=FALSE,results='hide',collapse=TRUE}

fit_data_filter = Maaslin2(
    input_data = df_input_data, 
    input_metadata = df_input_metadata, 
    normalization = "NONE",
    output = "demo_output_filter", 
    fixed_effects = c("diagnosis", "dysbiosis"),
    reference = c("diagnosis,nonIBD"),
    random_effects = c("subject"),
    min_prevalence = 0.1,
    min_abundance = 0.0001)

```